Fail Faster

### 2025-03-15 paper/paper.tex
wrote splitmix section

\subsection{Faster SplitMix with Unboxed Int}
\label{subsection:faster-rng}

\label{subsection:faster-rng}

@@

The precise details of how the SplitMix algorithm works \cn are unimportant for the
present paper, but the critical component is that all of its operations are are
defined in terms of arithmetic bitwise operations on 64-bit integers. In OCaml,
because of details related to the garbage collector, the 64-bit integer type \texttt{int64} is represented
at run time as a \emph{pointer} to an unscanned block of memory containing (among other things)
a 64-bit integer \cn.

This means that all operations that return an \texttt{int64} must allocate this block of memory.
This has a significant impact on the performance of generators. A single call to
one of the \bq library functions --- like generating an integer uniformly in a range --- may call the \texttt{splittable\_random}
algorithm multiple times. Each sample from \texttt{splittable\_random} allocates 9
words \jwc{this is a call to \texttt{next\_int64}}, and each allocation brings us closer
to the next garbage collection pause. While allocation in OCaml is fast and the garbage colletor
runs incrementally in small slices
\footnote{The OCaml GC is a generational collector \cn, and since these allocations are mostly very short lived, they will never be promoted. }, we will see in Section~\ref{section:eval} that this can have
a large performance impact on some generators that spend most of their time sampling data.

% \jwc{
%   \begin{itemize}
%     \item as we discussed in section~\ref{section:motiv}, 
%     \item faster rngs and sources of entropy exist, though they are extensionally inequivalent. 
%     \item here's how splitmix is slow: the core of splitmix is nextint64, which does allocations for each operation.
%     \item if we could use actual int64s, we wouldn't have to allocate.
%     \item so we do this:
%     \begin{itemize}
%       \item we implement splitmix in c (which uses normal int64s) and real int64 arithmetic, and use the ocaml ffi to call out to the c functions. t
%       \item this could be done with jane street bleeding-edge compiler extensions, though these are not yet implemented in a a version of the ocaml compiler that also has staging.
%     \end{itemize}
%   \end{itemize}
% }

>>>>

\subsection{Faster SplitMix with Unboxed \texttt{int64}s}
\label{subsection:faster-rng}

@@

The precise details of how the SplitMix algorithm works \cn are unimportant for the
present paper, but the critical component is that all of its operations are are
defined in terms of arithmetic bitwise operations on 64-bit integers. In OCaml,
because of details related to the garbage collector, the 64-bit integer type \texttt{int64} is represented
at run time as a \emph{pointer} to an unscanned block of memory containing (among other things)
a 64-bit integer \cn. This means that all operations that return an \texttt{int64} must allocate this block of memory.
This has a significant impact on the performance of generators. A single call to
one of the \bq library functions --- like generating an integer uniformly in a range --- may call the \texttt{splittable\_random}
algorithm multiple times. Each sample from \texttt{splittable\_random} allocates 9
words \jwc{this is a call to \texttt{next\_int64}}, and each allocation brings us closer
to the next garbage collection pause. While allocation in OCaml is fast and the garbage colletor
runs incrementally in small slices
\footnote{The OCaml GC is a generational collector \cn, and since these allocations are mostly very short lived, they will never be promoted. }, we will see in Section~\ref{section:eval} that this can have
a large performance impact on some generators that spend most of their time sampling data.
To circumvent this allocation and provide an equivalent version of \texttt{splittable\_random}, we
reimplement SplitMix in C, and call out to it with the OCaml FFI. The C version of the library uses
proper \texttt{int64\_t} arithmetic, only boxing and unboxing integers at the call boundaries between OCaml and C code.
Ideally in the future, one would not need to call out to C for this: the Jane Street bleeding-edge OCaml compiler has support
for unboxed types \cn, which (among other things) would let us implement a version of SplitMix that does not allocate, directly
in OCaml. Unfortunately, the Jane Street branch of the compiler is incompatible with MetaOCaml, which we use to implement the metaprogramming
discussed in the previous sections.
